New cardinality estimation algorithms for HyperLogLog sketches

### 2016-12-03 paper/paper.tex
some minor changes
added chart showing the probability of equal registers as a function of the Jaccard distance + some minor improvements

\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symImaginary}{\rm{i}}
\symDefine{\symIndexJ}{j}
\symDefine{\symIndexL}{l}
\symDefine{\symCount}{c}
\symDefine{\symCountVariate}{C}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symKappa}{\kappa}
\symDefine{\symX}{x}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symPhi}{\varphi}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symHelperW}{w}
\symDefine{\symHelperV}{v}
\symDefine{\symHelperU}{u}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}
\symDefine{\symBiasCorrectionFunc}{w_{\text{corr}}}

@@

Let us assume that the register values have following cumulative distribution function
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. For a random variable $\symRegValVariate$ obeying \eqref{equ:assumed_register_val_distribution} the expectation of $2^{-\symRegValVariate}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegValVariate})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
\left(
e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}
-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal-1}}}
\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
is a smooth periodic function with period 1. 
Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}). This value can also be found using Fourier analysis as shown in in \cref{app:analysis_xi}.

@@

Putting all these optimizations together finally gives the new cardinality estimation algorithm as presented in \cref{alg:estimate_ml}. The algorithm requires mainly only elementary operations. The square root that appears for the calculation of the stopping limit $\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$ can be precomputed, because it only depends on the HyperLogLog precision parameter $\symPrecision$. For large cardinalities it is worth to take the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as initial value, which requires a single logarithm evaluation. The stronger bound is a much better approximation especially for large cardinalities, where the extra logarithm evaluation is amortized by the savings of iteration cycles.

@@

\subsection{Performance}
We also measured the performance of \cref{alg:estimate_ml} using the same test setup as described in \cref{sec:corrected_raw_estimation_algorithm}. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}. The average computation time for the maximum likelihood algorithm shows a different behavior than for the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}). As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications.

@@

\begin{table}
\centering
\caption{List of true cardinalities of the pairwise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated. The last column shows the average number of log-likelihood function 
 \eqref{equ:max_likelihood_phi_func} and gradient \eqref{equ:df_dphi} evaluations until the stop criterion of the BFGS algorithm was satisfied.
}
\label{tbl:joint_estimation_cases}
\csvreader[before reading=\scriptsize, tabular=r|rrr|rr|rr, table head= & 
\multicolumn{3}{c|}{true cardinalities} &
\multicolumn{1}{c}{Jaccard index} &
\multicolumn{1}{c|}{cardinality ratio} &
\multicolumn{1}{c}{avg number}
\\
\# &  
\multicolumn{1}{c}{$|\symSetA|$} & 
\multicolumn{1}{c}{$|\symSetB|$} & 
\multicolumn{1}{c|}{$|\symSetX|$} & 
\multicolumn{1}{c}{$|\symSetS_1\cap\symSetS_2|/|\symSetS_1\cup\symSetS_2|$} & 
\multicolumn{1}{c|}{$|\symSetS_1|/|\symSetS_2|$} &
\multicolumn{1}{c}{iterations}
\\\hline]
{joint_cardinality_calculation.csv}{
trueCardA=\trueCardA,
trueCardB=\trueCardB,
trueCardX=\trueCardX,
trueJaccardIdx=\trueJaccardIdx,
trueRatio=\trueRatio,
avgNumEvaluations=\avgNumEvaluations
}{
\thecsvrow & 
\num{\trueCardA} & 
\num{\trueCardB} & 
\num{\trueCardX} & 
\numformat{\trueJaccardIdx} & 
\numformattwo{\trueRatio} &
\numformattwo{\avgNumEvaluations}
}
\end{table}

@@

\section{Future work}
As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased for the estimation of the cardinality. We have shown that this approach works well for the corrected raw estimator as well as the maximum likelihood estimator even though both are only approximately unbiased. Therefore, it would be interesting to investigate, which conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness, if used as cardinality estimator.

@@

The HyperLogLog algorithm itself can be regarded as hashing algorithm as it maps sets to register values. For sufficiently large cardinalities we can use the Poisson approximation and assume that the number of zero-valued HyperLogLog register can be ignored. Furthermore, if the HyperLogLog parameter $\symRegRange$ is chosen large enough, the number of saturated registers can be ignored as well. As a consequence, we can assume that the distribution of register values follows \eqref{equ:assumed_register_val_distribution} and the probability that a register has the same value for two different sets is (compare \eqref{equ:register_value_joint_pmf})
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
=
\sum_{\symRegVal = -\infty}^{\infty}
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\left(
1
-
e^{-\frac{\symPoissonRate_\symSetASuffix +  \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
+
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\right).
\end{equation}
Using the approximation 
$\sum_{\symRegVal=-\infty}^\infty
e^{-\frac{\symX}{2^\symRegVal}} - e^{-\frac{\symY}{2^\symRegVal}}
\approx
2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ we get
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\left(
\log\!\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetASuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
+
\log\!\left(
1
-
\frac{1}{2}
\frac{\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
\right)
\right).
\end{equation}
Further reduction by using the first order approximation $\log(1-\symX)\approx -\symX$ finally shows that the probability of a register having the same value for both sets is roughly linear to the Jaccard index $\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
\symAlpha_\infty
\frac{\symPoissonRate_\symSetXSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}
+
(1-\symAlpha_\infty).
\end{equation}
This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing. What also looks promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}). In contrast, min-wise hashing is very costly, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}.

@@

\section{Analysis of $\symPowerSeriesFunc(\symX)$}
\label{app:analysis_xi}
The Fourier series of the function $\symPowerSeriesFunc(\symX)$, which is defined by
\begin{equation}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
and which has period equal to 1, is
\begin{equation}
\symPowerSeriesFunc(\symX) = 1 + \Re\!\left(\sum_{\symIndexL=1}^{\infty}
\symA_\symIndexL \exp(2\pi\symIndexL\symX\symImaginary)\right)
\end{equation}
with coefficients
\begin{equation}
\symA_\symIndexL = 2\int_0^1 \symPowerSeriesFunc(\symX) \exp(-2\pi\symIndexL\symX\symImaginary)d\symX
=
2\log(2) \int_{-\infty}^{\infty} 2^\symX
e^{-2^\symX}\exp(-2\pi\symIndexL\symX\symImaginary)d\symX.
\end{equation}
The variable transformation $\symY = 2^\symX$ yields
\begin{equation}
\symA_\symIndexL
=
2\int_{0}^{\infty} e^{-\symY}
\symY^{-\frac{2\pi\symIndexL\symImaginary}{\log(2)}}d\symY
= 
2\Gamma\!\left(1-\frac{2\pi\symIndexL\symImaginary}{\log(2)}\right)
\end{equation}
where $\Gamma$ denotes the gamma function. Using the identity $\left|\Gamma(1+\symImaginary\symX)\right| = 
\sqrt{\frac{\pi\symX}{\sinh(\pi\symX)}}
$
we are able to write for the absolute values of the coefficients
\begin{equation}
|\symA_\symIndexL| = 
2\sqrt{\frac{\symB\symIndexL}{\sinh(\symB\symIndexL)}}
\quad
\text{with}
\
\symB:=\frac{2\pi^2}{\log 2}.
\end{equation}
In particular, the amplitude of the first harmonic is 
$|\symA_1|
\approx
\num{9.884e-6}$. Obviously, the maximum deviation of $\symPowerSeriesFunc(\symX)-1$ from the first harmonic must be smaller than $\sum_{\symIndexL=2}^{\infty} 
|\symA_\symIndexL|$. The ratio of subsequent coefficients is given by
\begin{equation}
\frac{|\symA_{\symIndexL+1}|}{|\symA_{\symIndexL}|}
=
\sqrt{\frac{\symIndexL+1}{\symIndexL}}
\sqrt{\frac{
\sinh(\symB \symIndexL)
}{
\sinh(\symB (\symIndexL+1))
}}
=
\sqrt{\frac{\symIndexL+1}{\symIndexL}}
\sqrt{\frac{
1
}{
\cosh(\symB)
+
\frac{
\sinh(\symB)
}
{
\tanh(\symB \symIndexL)
}
}}.
\end{equation}
For $\symIndexL\geq 2$ we have $\sqrt{\frac{\symIndexL+1}{\symIndexL}}\leq \sqrt{\frac{3}{2}}$. Together with $\tanh(x)\leq 1$ we obtain
\begin{equation}
\frac{|\symA_{\symIndexL+1}|}{
|\symA_{\symIndexL}|
}
\leq
\sqrt{\frac{3}{2}}
\sqrt{\frac{
1
}{
\cosh(\symB)
+
\sinh(\symB)
}}
\\
=
\sqrt{\frac{
3
}{
2 e^\symB
}}.
\end{equation}
Consequently, $|\symA_{\symIndexL}| \leq 
|\symA_{2}|
\left(\sqrt{\frac{
3
}{
2 e^\symB
}}\right)^{\symIndexL-2}$ for $\symIndexL\geq 2$ and 
\begin{equation}
\sum_{\symIndexL=2}^{\infty} 
|\symA_\symIndexL|
\leq
|\symA_2|
\sum_{\symIndexL=0}^{\infty} 
\left(
\sqrt{\frac{
3
}{
2e^\symB
}}
\right)^\symIndexL
=
2\sqrt{
\frac{2\symB}{\sinh(2\symB)}}
\frac{1}{1-
\sqrt{\frac{
3
}{
2 e^\symB
}}
}
\approx
\num{9.154e-12}.
\end{equation}
For the maximum deviation from 1 we finally have
\begin{equation}
\max(|\symPowerSeriesFunc(\symX)-1|) \leq \sum_{\symIndexL=1}^{\infty} 
|\symA_\symIndexL|
\leq \num{9.885e6}.
\end{equation}

>>>>

\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symDistanceMeasure}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symImaginary}{\rm{i}}
\symDefine{\symIndexJ}{j}
\symDefine{\symIndexL}{l}
\symDefine{\symCount}{c}
\symDefine{\symCountVariate}{C}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symKappa}{\kappa}
\symDefine{\symX}{x}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symPhi}{\varphi}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symHelperW}{w}
\symDefine{\symHelperV}{v}
\symDefine{\symHelperU}{u}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}
\symDefine{\symBiasCorrectionFunc}{w_{\text{corr}}}

@@

Let us assume that the register values have following cumulative distribution function
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. For a random variable $\symRegValVariate$ obeying \eqref{equ:assumed_register_val_distribution} the expectation of $2^{-\symRegValVariate}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegValVariate})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
\left(
e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}
-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal-1}}}
\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
is a smooth periodic function with period 1. 
Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}). This value can also be found using Fourier analysis as shown in \cref{app:analysis_xi}.

@@

Putting all these optimizations together finally gives the new cardinality estimation algorithm as presented in \cref{alg:estimate_ml}. The algorithm requires mainly only elementary operations. The square root that appears for the calculation of the stopping limit $\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$ can be precomputed, because it only depends on the HyperLogLog precision parameter $\symPrecision$. For large cardinalities it is worth to take the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as initial value, which requires a single logarithm evaluation. The stronger bound is a much better approximation especially for large cardinalities, where the extra logarithm evaluation is amortized by savings in the number of iteration cycles.

@@

\subsection{Performance}
We also measured the performance of \cref{alg:estimate_ml} using the same test setup as described in \cref{sec:corrected_raw_estimation_algorithm}. The results for HyperLogLog configurations $\symPrecision=12, \symRegRange=20$ and $\symPrecision=12, \symRegRange=52$ are shown in \cref{fig:avg_exec_time}. The average computation time for the maximum likelihood algorithm shows a different behavior than for the corrected raw estimation algorithm (compare \cref{fig:corrected_raw_avg_exec_time}).As can be seen, the average execution time is larger for most cardinalities, but nevertheless, it never exceeds \SI{700}{\nano\second} which is still fast enough for many applications. The steps in the chart can be explained by different numbers of iteration cycles until the secant method is stopped. For example, at a cardinality value around \num{5000} the average number of required cycles until the stop criterion is met increases abruptly from two to three. More than three iteration cycles have never been necessary for any cardinality estimate in this performance test.

@@

\begin{table}
\centering
\caption{List of true cardinalities of the pairwise disjoint sets $\symSetA$, $\symSetB$, and $\symSetX$, for which maximum likelihood estimation from two HyperLogLog sketches with parameters $\symPrecision=20$ and $\symRegRange=44$ representing sets $\symSetS_1 = \symSetA\cup\symSetX$ and  $\symSetS_2 = \symSetB\cup\symSetX$ was investigated. The last column shows the average number of iteration cycles until the stop criterion of the BFGS algorithm was satisfied.
}
\label{tbl:joint_estimation_cases}
\csvreader[before reading=\scriptsize, tabular=r|rrr|rr|rr, table head= & 
\multicolumn{3}{c|}{true cardinalities} &
\multicolumn{1}{c}{Jaccard index} &
\multicolumn{1}{c|}{cardinality ratio} &
\multicolumn{1}{c}{avg number}
\\
\# &  
\multicolumn{1}{c}{$|\symSetA|$} & 
\multicolumn{1}{c}{$|\symSetB|$} & 
\multicolumn{1}{c|}{$|\symSetX|$} & 
\multicolumn{1}{c}{$|\symSetS_1\cap\symSetS_2|/|\symSetS_1\cup\symSetS_2|$} & 
\multicolumn{1}{c|}{$|\symSetS_1|/|\symSetS_2|$} &
\multicolumn{1}{c}{iterations}
\\\hline]
{joint_cardinality_calculation.csv}{
trueCardA=\trueCardA,
trueCardB=\trueCardB,
trueCardX=\trueCardX,
trueJaccardIdx=\trueJaccardIdx,
trueRatio=\trueRatio,
avgNumEvaluations=\avgNumEvaluations
}{
\thecsvrow & 
\num{\trueCardA} & 
\num{\trueCardB} & 
\num{\trueCardX} & 
\numformat{\trueJaccardIdx} & 
\numformattwo{\trueRatio} &
\numformattwo{\avgNumEvaluations}
}
\end{table}

@@

\section{Future work}
As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased for the estimation of the cardinality. We have shown that this approach works well for the corrected raw estimator as well as for the maximum likelihood estimator even though both are only approximately unbiased. Therefore, it would be interesting what conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness, if used as cardinality estimator.

@@

The HyperLogLog algorithm itself can be regarded as hashing algorithm as it maps sets to register values. For sufficiently large cardinalities we can use the Poisson approximation and assume that the number of zero-valued HyperLogLog registers can be ignored. Furthermore, if the HyperLogLog parameter $\symRegRange$ is chosen large enough, the number of saturated registers can be ignored as well. As a consequence, we can assume that the distribution of register values follows \eqref{equ:assumed_register_val_distribution} and the probability that a register has the same value for two different sets is (compare \eqref{equ:register_value_joint_pmf})
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
=
\sum_{\symRegVal = -\infty}^{\infty}
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\left(
1
-
e^{-\frac{\symPoissonRate_\symSetASuffix +  \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
-
e^{-\frac{\symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
+
e^{-\frac{\symPoissonRate_\symSetASuffix + \symPoissonRate_\symSetBSuffix + \symPoissonRate_\symSetXSuffix}
{\symNumReg 2^\symRegVal}
}
\right).
\end{equation}
Using the approximation 
$\sum_{\symRegVal=-\infty}^\infty
e^{-\frac{\symX}{2^\symRegVal}} - e^{-\frac{\symY}{2^\symRegVal}}
\approx
2\symAlpha_\infty \left(\log(\symY)-\log(\symX)\right)$ we get

\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\frac{1}{2}
\symDistanceMeasure
+
\frac{1}{4}
\symDistanceMeasure^2
\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
\right)
\end{equation}
where $\symDistanceMeasure = \frac{\symPoissonRate_\symSetASuffix
+\symPoissonRate_\symSetBSuffix}
{\symPoissonRate_\symSetASuffix+\symPoissonRate_\symSetBSuffix+
\symPoissonRate_\symSetXSuffix}$
 is the Jaccard distance. Since $
\frac{
\symPoissonRate_\symSetASuffix
\symPoissonRate_\symSetBSuffix
}{\left(
\symPoissonRate_\symSetASuffix+
\symPoissonRate_\symSetBSuffix
\right)^2}
$ is always in the range $[0,\frac{1}{4}]$, the probability for equal register values can be bounded by
\begin{equation}
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\frac{1}{2}
\symDistanceMeasure
\right)
\lesssim
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\lesssim
1
+
2\symAlpha_\infty
\log\!\left(
1
-
\frac{1}{2}
\symDistanceMeasure
+
\frac{1}{16}
\symDistanceMeasure^2
\right).
\end{equation}
As shown in \cref{fig:equal_register_probability} the bounds are very close, especially for small Jaccard distances where the probability can be well approximated by
\begin{equation}
\symProbability(\symRegValVariate_1=\symRegValVariate_2)
\approx
1-
\symAlpha_\infty\symDistanceMeasure.
\end{equation}
\begin{figure}
\centering
\includesvg[width=0.6\textwidth]{equal-register-probability}
\caption{The approximate probability range of equal register values as a function of the Jaccard distance.}
\label{fig:equal_register_probability}
\end{figure}
This finding makes the HyperLogLog algorithm an interesting candidate for locality-sensitive hashing. What also looks promising is the much simpler preprocessing because it only involves a single hash function (compare \cref{alg:insert}). In contrast, min-wise hashing is very costly, because many hash function evaluations are necessary for all the needed permutations \cite{Li2011}.

@@

\section{Analysis of $\symPowerSeriesFunc(\symX)$}
\label{app:analysis_xi}
The Fourier series of the function $\symPowerSeriesFunc(\symX)$, which is defined by
\begin{equation}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
and which has period equal to 1, is
\begin{equation}
\symPowerSeriesFunc(\symX) = 1 + \Re\!\left(\sum_{\symIndexL=1}^{\infty}
\symA_\symIndexL e^{2\pi\symImaginary\symIndexL\symX}\right)
\end{equation}
with coefficients
\begin{multline}
\symA_\symIndexL = 2\int_0^1 \symPowerSeriesFunc(\symX) e^{-2\pi\symImaginary\symIndexL\symX}d\symX
=
2\log(2)
\int_0^1
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
e^{-2\pi\symImaginary\symIndexL\symX}
d\symX
=
\\
=
2\log(2)
\sum_{\symRegVal = -\infty}^\infty
\int_{\symRegVal}^{\symRegVal+1}
2^{\symX}
e^{-2^{\symX}}
e^{-2\pi\symImaginary\symIndexL\symX}
d\symX
=
2\log(2) \int_{-\infty}^{\infty} 2^\symX
e^{-2^\symX}
e^{-2\pi\symImaginary\symIndexL\symX}d\symX.
\end{multline}
The variable transformation $\symY = 2^\symX$ yields
\begin{equation}
\symA_\symIndexL
=
2\int_{0}^{\infty} e^{-\symY}
\symY^{-\frac{2\pi\symImaginary\symIndexL}{\log(2)}}d\symY
= 
2\Gamma\!\left(1-\frac{2\pi\symImaginary\symIndexL}{\log(2)}\right)
\end{equation}
where $\Gamma$ denotes the gamma function. Using the identity $\left|\Gamma(1+\symImaginary\symX)\right| = 
\sqrt{\frac{\pi\symX}{\sinh(\pi\symX)}}
$
we are able to write for the absolute values of the coefficients
\begin{equation}
|\symA_\symIndexL| = 
2\sqrt{\frac{\symB\symIndexL}{\sinh(\symB\symIndexL)}}
\quad
\text{with}
\
\symB:=\frac{2\pi^2}{\log 2}.
\end{equation}
In particular, the amplitude of the first harmonic is 
$|\symA_1|
\approx
\num{9.884e-6}$. Obviously, the maximum deviation of $\symPowerSeriesFunc(\symX)-1$ from the first harmonic must be smaller than $\sum_{\symIndexL=2}^{\infty} 
|\symA_\symIndexL|$. The ratio of subsequent coefficients is given by
\begin{equation}
\frac{|\symA_{\symIndexL+1}|}{|\symA_{\symIndexL}|}
=
\sqrt{\frac{\symIndexL+1}{\symIndexL}}
\sqrt{\frac{
\sinh(\symB \symIndexL)
}{
\sinh(\symB (\symIndexL+1))
}}
=
\sqrt{\frac{\symIndexL+1}{\symIndexL}}
\sqrt{\frac{
1
}{
\cosh(\symB)
+
\frac{
\sinh(\symB)
}
{
\tanh(\symB \symIndexL)
}
}}.
\end{equation}
For $\symIndexL\geq 2$ we have $\sqrt{\frac{\symIndexL+1}{\symIndexL}}\leq \sqrt{\frac{3}{2}}$. Together with $\tanh(x)\leq 1$ we obtain
\begin{equation}
\frac{|\symA_{\symIndexL+1}|}{
|\symA_{\symIndexL}|
}
\leq
\sqrt{\frac{3}{2}}
\sqrt{\frac{
1
}{
\cosh(\symB)
+
\sinh(\symB)
}}
\\
=
\sqrt{\frac{
3
}{
2 e^\symB
}}.
\end{equation}
Consequently, $|\symA_{\symIndexL}| \leq 
|\symA_{2}|
\left(\sqrt{\frac{
3
}{
2 e^\symB
}}\right)^{\symIndexL-2}$ for $\symIndexL\geq 2$ and 
\begin{equation}
\sum_{\symIndexL=2}^{\infty} 
|\symA_\symIndexL|
\leq
|\symA_2|
\sum_{\symIndexL=0}^{\infty} 
\left(
\sqrt{\frac{
3
}{
2e^\symB
}}
\right)^\symIndexL
=
2\sqrt{
\frac{2\symB}{\sinh(2\symB)}}
\frac{1}{1-
\sqrt{\frac{
3
}{
2 e^\symB
}}
}
\approx
\num{9.154e-12}.
\end{equation}
For the maximum deviation from 1 we finally have
\begin{equation}
\max(|\symPowerSeriesFunc(\symX)-1|) \leq \sum_{\symIndexL=1}^{\infty} 
|\symA_\symIndexL|
\leq \num{9.885e-6}.
\end{equation}
